Geographically aware estimates of remotely sensed water properties for Chesapeake Bay

### 2022-09-20 manuscript/manuscript.tex
text edits

We compiled over two decades of observational data (2000-2021) on salinity (n=6k), temperature (n=10k), and turbidity (n=5k, Table S1). Although this data came from a wide variety of sources (n=20, Table S1), almost $80\%$ of the total unique observations came from three programs NDBC/CBIBS (\url{https://buoybay.noaa.gov/}), Maryland Continuous Monitoring Program (MDNR), and the Virginia Estuarine and Coastal Observing System (\url{http://vecos.vims.edu}). Because Chesapeake Bay has a mean optical depth of 0.89m \citep{urquhartRemotelySensedEstimates2012}, we filtered all collected observations to those occurring in the top 1m of the water column. We further limited observations to those that had a timestamp during the MODIS overpass window time-of-day (approximately 1700 - 1900 hours). Prior to model fitting, we removed the annual cycle from temperature observations using least squares optimization routines \citep{virtanenSciPyFundamentalAlgorithms2020a}. Finally, we implemented spatial and temporal averaging of observational data to a daily mean within each MODIS pixel.

@@

Our tuned RF models for salinity, turbidity, and temperature had a root mean square error (RMSE) of 2.34 ($R^2$=0.91), 32.2 ($R^2$=0.07), and 1.8 ($R^2$=0.96) respectively (Table 1, Figure 1). Out of sample validation performance was excellent for salinity and temperature models but poor for the turbidity model (Figure 1, Figure S1). Despite excellent overall performance, our temperature and salinity models had poor performance at the extremes. They were unable to recover very low temperature values ($<$ 0 C) or very high temperature and salinity values ($>$ 30 C, $>$ 23 salinity, Figure 1). One explanation for these areas of poor performance are their relative rarity in the observational dataset (Table S2). The salinity model however did faithfully capture areas of low salinity (Figure 1). These low salinity values can be regarded as low uncertainty (high confidence) estimates given the extent to which the upper tributaries are well represented in the observational dataset (Figure S2).

@@

\begin{figure}[ht!]
    \begin{center}
          \includegraphics[width=0.75\textwidth,keepaspectratio]{figures/_seasonality}
          \caption{Seasonal Random Forest prediction results for temperature and salinity.}
    \end{center}    
\end{figure}

@@

Our geographically aware approach was able to accurately reproduce spatial and temporal patterns of Bay salinity and temperature. Model errors were generally lower than prior data-driven efforts and as good or better than physics-based models \citep{vogelAssessingSatelliteSea2016}. In addition to being accurate, our approach provides the ability to generate rapid bay-wide prediction surfaces in seconds and it is easily updatable insofar as saved models can be reloaded and have their training “updated”. Our models are lightweight with minimal internal complexity and have been extensively tested against a uniquely large observational dataset.

@@

Often it is difficult to assess the inherent uncertainty of data-driven efforts such as ours. However, in our case, we assembled such a large observational dataset, that it makes such assessments possible. This is a key difference between our approach and prior efforts which have mostly been trained on high salinity buoy data or in-situ cruises from the mainstem of the Bay \citep{vogelAssessingSatelliteSea2016, geigerSatellitederivedCoastalOcean2013}. Predicted areas of low salinity in particular likely have low associated uncertainties because the upper tributaries were represented with high density in our observational dataset (Figure S2). Because CBOFS salinity data was fresher than our RF estimates in the upper tributaries, this suggests that CBOFS data for these areas are true underestimates and not a deficiency of the RF model.

>>>>

We compiled over two decades of observational data (2000-2021) on salinity (n=6k), temperature (n=10k), and turbidity (n=5k, Table S1). Although this data came from a wide variety of sources (n=20, Table S1), almost $80\%$ of the total unique observations came from three programs NDBC/CBIBS (\url{https://buoybay.noaa.gov/}), Maryland Continuous Monitoring Program (MDNR), and the Virginia Estuarine and Coastal Observing System (\url{http://vecos.vims.edu}). The temporal range of the observations includes both times of minimum and maximum freshwater input to the Bay (\url{https://www.usgs.gov/media/images/estimated-annual-mean-streamflow-entering-chesapeake-bay}).

Because Chesapeake Bay has a mean optical depth of 0.89m \citep{urquhartRemotelySensedEstimates2012}, we filtered all collected observations to those occurring in the top 1m of the water column. We further limited observations to those that had a timestamp during the MODIS overpass window time-of-day (approximately 1700 - 1900 hours). Prior to model fitting, we removed the annual cycle from temperature observations using least squares optimization routines \citep{virtanenSciPyFundamentalAlgorithms2020a}. Finally, we implemented spatial and temporal averaging of observational data to a daily mean within each MODIS pixel.

@@

Our tuned RF models for salinity, turbidity, and temperature had a root mean square error (RMSE) of 2.34 ($R^2$=0.91), 32.2 ($R^2$=0.07), and 1.8 ($R^2$=0.96) respectively (Table 1, Figure 1). Out of sample validation performance was excellent for salinity and temperature models but poor for the turbidity model (Figure 1, Figure S1). Despite excellent overall performance for the temperature and salinity models, they had poor performance at the extremes. They were unable to recover very low temperature values ($<$ 0 C) or very high temperature and salinity values ($>$ 30 C, $>$ 23 salinity, Figure 1). One explanation for these areas of poor performance are their relative rarity in the observational dataset (Table S2). The salinity model however did faithfully capture areas of low salinity (Figure 1). These low salinity values can be regarded as low uncertainty (high confidence) estimates given the extent to which the upper tributaries are well represented in the observational dataset (Figure S2).

@@

\begin{figure}[ht!]
    \begin{center}
          \includegraphics[width=0.7\textwidth,keepaspectratio]{figures/_seasonality}
          \caption{Seasonal Random Forest prediction results for temperature and salinity.}
    \end{center}    
\end{figure}

@@

Our geographically aware approach was able to accurately reproduce realistic spatial and temporal patterns of Bay salinity and temperature. Model errors were generally lower than prior data-driven efforts and as good or better than physics-based models \citep{vogelAssessingSatelliteSea2016}. In addition to being accurate, our approach provides the ability to generate rapid bay-wide prediction surfaces in seconds and it is easily updatable insofar as saved models can be reloaded and have their training “updated”. Our models are lightweight with minimal internal complexity and have been extensively tested against a uniquely large observational dataset.

One reason why the turbidity model may have had poor performance is because our observational data only contains data in Nephelometric Turbidity Units (NTUs) which are not standardized units. As a result, they cannot be reliably compared across different instruments. A better turbidity model would likely result from using total suspended matter (TSM) data measured in a standardized unit such as mg/l \citep{ondrusekDevelopmentNewOptical2012}.

@@

Often it is difficult to assess the inherent uncertainty of data-driven efforts such as ours. However, in our case, we assembled such a large observational dataset, that it makes such assessments possible. This is a key difference between our approach and prior efforts which have mostly been trained on high salinity buoy data or in-situ cruises from the mainstem of the Bay \citep{vogelAssessingSatelliteSea2016, geigerSatellitederivedCoastalOcean2013, ondrusekDevelopmentNewOptical2012}. Predicted areas of low salinity in particular likely have low associated uncertainties because the upper tributaries were represented with high density in our observational dataset (Figure S2). Because CBOFS salinity data was fresher than our RF estimates in the upper tributaries, this suggests that CBOFS data for these areas are true underestimates and not a deficiency of the RF model.

In contrast to the upper tributaries, an area of maximum uncertainty is likely the mainstem and eastern Bay across from the mouth of the Potomac (Tangier Sound). This appears to be the upper limit of the marine plume emanating from the Bay mouth. The hydrodynamics here seem to be complex given that the CBOFS has difficulty resolving spatial patterning in this area. Improving our data-driven modeling approach would be difficult here because this area has few observational data (Figure S2).
